Generalized Rybicki Press algorithm

### 2014-09-30 paper/ESS.tex
Fixed corrections

To analyze the scaling with respect to $p$ recall that the computational cost of solving a banded matrix of size $M \times M$, whose bandwidth is $B$ scales as $\mathcal{O}(B^2M)$. Hence, in our case, a naive complexity analysis shows that the scaling of the algorithm is $\mathcal{O}(p^3N)$. However, note that within the band

@@

The extended system, denoted by $A_{ex}$ on appropriate reordering of rows and columns can be written as
\begin{align}
P_1A_{ex} P_2
\begin{bmatrix}
l_1^{(p)}\\
l_2^{(p)}\\
\vdots\\
l_{N-1}^{(p)}\\
r_2^{(p)}\\
r_3^{(p)}\\
\vdots\\
r_N^{(p)}\\
x_1\\
x_2\\
\vdots\\
x_N
\end{bmatrix} & =
\begin{bmatrix}
I_1 & 0 & U_a\\
0 & I_2 & V_a\\
V_b & U_b & D
\end{bmatrix}
\begin{bmatrix}
l_1^{(p)}\\
l_2^{(p)}\\
\vdots\\
l_{N-1}^{(p)}\\
r_2^{(p)}\\
r_3^{(p)}\\
\vdots\\
r_N^{(p)}\\
x_1\\
x_2\\
\vdots\\
x_N
\end{bmatrix}
\end{align}
where $P_1$, $P_2$ are permutation matrices, the matrix $I_1$ is a highly sparse lower-triangular matrix with $1$'s on the diagonal and $-1$'s at a few places in the lower-triangular part (the precise location is important for determinant computations as we will see later), the matrix $I_2$ is a highly sparse upper-triangular matrix with $1$'s on the diagonal and $-1$'s at a few places in the upper-triangular part and $D$ is a diagonal matrix. The first set of rows, i.e., $\begin{bmatrix} I_1 & 0 & U_a\end{bmatrix}$, correspond to adding the variables $l_k^{(p)}$, i.e., $l_k^{(p)} = U_k^Tx_k + l_{k-1}^{(p)}$. The next set of rows, i.e., $\begin{bmatrix} 0 & I_2 & V_a\end{bmatrix}$, correspond to adding the variables $r_k^{(p)}$, i.e., $r_k^{(p)} = V_k^Tx_k + r_{k+1}^{(p)}$. The last set of rows, i.e., $\begin{bmatrix} V_b & U_b & D\end{bmatrix}$, correspond to the initial set of equations with the $l_k^{(p)}$'s and $r_k^{(p)}$'s introduced.
We then have
\begin{align}
\det(P_1A_{ex}P_2) & =
\underbrace{
\det\left(
\begin{bmatrix}
I_1 & 0 & U_a\\
0 & I_2 & V_a\\
V_b & U_b & D
\end{bmatrix}
\right) =
\det\left(
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}
\right)
\det\left(D-
\begin{bmatrix}
U_a\\
V_a
\end{bmatrix}
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}^{-1}
\begin{bmatrix}
U_b & V_b
\end{bmatrix}
\right)
}_{\text{Block determinant formula}}
\end{align}
Now note that $\det(I_1) = 1 = \det(I_2)$, due to the fact that $I_1$ and $I_2$ are triangular matrices with $1$'s on the diagonal. Hence,
\begin{align}
\det\left(
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}
\right) & = \det(I_1) \det(I_2) = 1 \times 1 = 1
\end{align}
Further, note that the matrix $D-
\begin{bmatrix}
U_a\\
V_a
\end{bmatrix}
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}^{-1}
\begin{bmatrix}
U_b & V_b
\end{bmatrix}$ is the Schur complement obtained by eliminating the variables $l_i^{(p)}$, $r_i^{(p)}$ and hence is the initial dense matrix $A$ we began with. Hence, we have
\begin{align}
\det(A_{ex}) & = \pm \det(A)
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%										%
%		SECTION: RYBICKI PRESS			%
%										%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reinterpretation of Rybicki Press algorithm in terms of sparse embedding}

We will first naively reinterpret the Rybicki Press algorithm in terms of the extended sparse matrix algebra. Recall that the Rybicki Press algorithm inverts a correlation matrix $A$ given by Equation~\eqref{eqn_Rybicki_Press_matrix}.
\begin{align}
A(i,j) & = \exp\left(-\beta \lvert t_i - t_j \rvert\right)
\label{eqn_Rybicki_Press_matrix}
\end{align}
where $t_i$'s lies on an interval and are monotone. The original Rybicki Press algorithm relies on the fact that the inverse of $A$ happens to be a tridiagonal matrix. The key ingredient of their algorithm is the following property of exponentials:
\begin{align}
\exp\left(\beta (t_i-t_j) \right)\exp\left(\beta (t_j-t_k) \right) = \exp\left(\beta (t_i-t_k)\right)
\end{align}
In our sparse interpretation as well, we will use this property to recognize that the matrix $A$ is a semi-separable matrix, whose semi-separable rank is $1$. This can be seen by setting $u_k = \exp(\beta t_k)$ and $v_k = \exp(-\beta t_k)$. This then gives us ($i<j$) that $A(i,j) = u_i v_j = \exp(\beta t_i) \exp(-\beta t_j) = \exp(\beta(t_i-t_j))$ and similarly for $i>j$. This shows that the matrix $A$ is semi-separable with semi-separable rank $1$. Hence, we can mimic the same approach as in the earlier sections to obtain an $\mathcal{O}(N)$ algorithm. However, there is an issue that needs to be addressed from a numerical perspective. If the $t_i$'s are spread over a large interval, then $u_i$ is exponentially large, while $v_i$ is exponentially small, and hence embedding into a sparse matrix as such could prove to be a catastrophic leading to underflow and overflow of the relevant entries. This issue though can be circumvented by a suitable analytic preconditioning, by an appropriate change of variables. This is illustrated for a $4 \times 4$ linear system. We will use the notation $t_{ij}$ to denote $\lvert t_i - t_j\lvert$. The linear equation is
\begin{align}
\begin{bmatrix}
1 & \exp(-\beta t_{12}) & \exp(-\beta t_{13}) & \exp(-\beta t_{14})\\
\exp(-\beta t_{12}) & 1 & \exp(-\beta t_{23}) & \exp(-\beta t_{24})\\
\exp(-\beta t_{13}) & \exp(-\beta t_{23}) & 1 & \exp(-\beta t_{34})\\
\exp(-\beta t_{14}) & \exp(-\beta t_{24}) & \exp(-\beta t_{34}) & 1\\
\end{bmatrix}
\begin{bmatrix}
x_1\\x_2\\x_3\\x_4
\end{bmatrix}
=
\begin{bmatrix}
b_1\\b_2\\b_3\\b_4
\end{bmatrix}
\end{align}
Now lets introduce the additional variables as follows:
\begin{align}
r_4 & = x_4\\
r_3 & = x_3 + \exp(-\beta t_{34})r_4\\
r_2 & = x_2 + \exp(-\beta t_{23})r_3
\end{align}
\begin{align}
l_2 & = x_1\exp(-\beta t_{12})\\
l_3 & = (x_2+l_2)\exp(-\beta t_{23})\\
l_4 & = (x_3+l_3)\exp(-\beta t_{34})
\end{align}
The equations then become
\begin{align}
x_1 + \exp(-\beta t_{12}) r_2 & = b_1\\
l_2 + x_2 + \exp(-\beta t_{23}) r_3 & = b_2\\
l_3 + x_3 + \exp(-\beta t_{24}) r_4 & = b_3\\
l_4 + x_4 & = b_4
\end{align}
Embedding this in an extended sparse matrix, we obtain
\begin{align}
\begin{bmatrix}
1 & \exp(-\beta t_{12}) & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
\exp(-\beta t_{12}) & 0 & -1 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
0 & -1 & 0 & 1 & \exp(-\beta t_{23}) & 0 & 0 & 0 & 0 & 0\\
0 & 0 & 1 & 1 & \exp(-\beta t_{23}) & 0 & 0 & 0 & 0 & 0\\
0 & 0 & \exp(-\beta t_{23}) & \exp(-\beta t_{23}) & 0 & -1 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & -1 & 0 & 1 & \exp(-\beta t_{34}) & 0 & 0\\
0 & 0 & 0 & 0 & 0 & 1 & 1 & \exp(-\beta t_{34}) & 0 & 0\\
0 & 0 & 0 & 0 & 0 & \exp(-\beta t_{34}) & \exp(-\beta t_{34}) & 0 & -1 & 0\\
0 & 0 & 0 & 0 & 0 & 0 & 0 & -1 & 0 & 1\\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
\end{bmatrix}
\begin{bmatrix}
x_1\\ r_2\\ l_1\\ x_2\\ r_3\\ l_2\\ x_3\\ r_4\\ l_3\\ x_4
\end{bmatrix}
=
\begin{bmatrix}
b_1\\ 0\\ 0\\ b_2\\ 0\\ 0\\ b_3\\ 0\\ 0\\ b_4
\end{bmatrix}
\end{align}

@@

\newpage
% \bibliography{biblio}
% \bibliographystyle{unsrt}
\begin{thebibliography}{10}

@@

>>>>

% To analyze the scaling with respect to $p$ recall that the computational cost of solving a banded matrix of size $M \times M$, whose bandwidth is $B$ scales as $\mathcal{O}(B^2M)$. Hence, in our case, a naive complexity analysis shows that the scaling of the algorithm is $\mathcal{O}(p^3N)$. However, note that within the band

@@

The extended system, denoted by $A_{ex}$ on appropriate reordering of rows and columns can be written as
\begin{align}
P_1A_{ex} P_2
\begin{bmatrix}
l_1^{(p)}\\
l_2^{(p)}\\
\vdots\\
l_{N-1}^{(p)}\\
r_2^{(p)}\\
r_3^{(p)}\\
\vdots\\
r_N^{(p)}\\
x_1\\
x_2\\
\vdots\\
x_N
\end{bmatrix} & =
\begin{bmatrix}
I_1 & 0 & U_a\\
0 & I_2 & V_a\\
V_b & U_b & D
\end{bmatrix}
\begin{bmatrix}
l_1^{(p)}\\
l_2^{(p)}\\
\vdots\\
l_{N-1}^{(p)}\\
r_2^{(p)}\\
r_3^{(p)}\\
\vdots\\
r_N^{(p)}\\
x_1\\
x_2\\
\vdots\\
x_N
\end{bmatrix}
\end{align}
where $P_1$, $P_2$ are permutation matrices, the matrix $I_1$ is a highly sparse lower-triangular matrix with $1$'s on the diagonal and $-1$'s at a few places in the lower-triangular part (the precise location is unimportant for determinant computations as we will see later), the matrix $I_2$ is a highly sparse upper-triangular matrix with $1$'s on the diagonal and $-1$'s at a few places in the upper-triangular part and $D$ is a diagonal matrix. The first set of rows, i.e., $\begin{bmatrix} I_1 & 0 & U_a\end{bmatrix}$, correspond to adding the variables $l_k^{(p)}$, i.e., $l_k^{(p)} = U_k^Tx_k + l_{k-1}^{(p)}$. The next set of rows, i.e., $\begin{bmatrix} 0 & I_2 & V_a\end{bmatrix}$, correspond to adding the variables $r_k^{(p)}$, i.e., $r_k^{(p)} = V_k^Tx_k + r_{k+1}^{(p)}$. The last set of rows, i.e., $\begin{bmatrix} V_b & U_b & D\end{bmatrix}$, correspond to the initial set of equations with the $l_k^{(p)}$'s and $r_k^{(p)}$'s introduced.
We then have
\begin{align}
\det(P_1A_{ex}P_2) & =
\underbrace{
\det\left(
\begin{bmatrix}
I_1 & 0 & U_a\\
0 & I_2 & V_a\\
V_b & U_b & D
\end{bmatrix}
\right) =
\det\left(
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}
\right)
\det\left(D-
\begin{bmatrix}
U_a\\
V_a
\end{bmatrix}
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}^{-1}
\begin{bmatrix}
U_b & V_b
\end{bmatrix}
\right)
}_{\text{Block determinant formula}}
\end{align}
Now note that $\det(I_1) = 1 = \det(I_2)$, due to the fact that $I_1$ and $I_2$ are triangular matrices with $1$'s on the diagonal. Hence,
\begin{align}
\det\left(
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}
\right) & = \det(I_1) \det(I_2) = 1 \times 1 = 1
\end{align}
Further, note that the matrix $D-
\begin{bmatrix}
U_a\\
V_a
\end{bmatrix}
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}^{-1}
\begin{bmatrix}
U_b & V_b
\end{bmatrix}$ is the Schur complement obtained by eliminating the variables $l_i^{(p)}$, $r_i^{(p)}$ and hence is the initial dense matrix $A$ we began with. Hence, we have
\begin{align}
\det(A_{ex}) & = \pm \det(A)
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%										%
%		SECTION: RYBICKI PRESS			%
%										%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reinterpretation of Rybicki Press algorithm in terms of sparse embedding}

We will first naively reinterpret the Rybicki Press algorithm in terms of the extended sparse matrix algebra. Recall that the Rybicki Press algorithm~\cite{rybicki1995class} inverts a correlation matrix $A$ given by Equation~\eqref{eqn_Rybicki_Press_matrix}.
\begin{align}
A(i,j) & = \exp\left(-\beta \lvert t_i - t_j \rvert\right)
\label{eqn_Rybicki_Press_matrix}
\end{align}
where $t_i$'s lies on an interval and are monotone. The original Rybicki Press algorithm relies on the fact that the inverse of $A$ happens to be a tridiagonal matrix. The key ingredient of their algorithm is the following property of exponentials:
\begin{align}
\exp\left(\beta (t_i-t_j) \right)\exp\left(\beta (t_j-t_k) \right) = \exp\left(\beta (t_i-t_k)\right)
\end{align}
In our sparse interpretation as well, we will use this property to recognize that the matrix $A$ is a semi-separable matrix, whose semi-separable rank is $1$. This can be seen by setting $u_k = \exp(\beta t_k)$ and $v_k = \exp(-\beta t_k)$. This then gives us ($i<j$) that $A(i,j) = u_i v_j = \exp(\beta t_i) \exp(-\beta t_j) = \exp(\beta(t_i-t_j))$ and similarly for $i>j$. This shows that the matrix $A$ is semi-separable with semi-separable rank $1$. Hence, we can mimic the same approach as in the earlier sections to obtain an $\mathcal{O}(N)$ algorithm. However, there is an issue that needs to be addressed from a numerical perspective. If the $t_i$'s are spread over a large interval, then $u_i$ is exponentially large, while $v_i$ is exponentially small, and hence embedding into a sparse matrix as such could prove to be a catastrophic leading to underflow and overflow of the relevant entries. This issue though can be circumvented by a suitable analytic preconditioning, by an appropriate change of variables. This is illustrated for a $4 \times 4$ linear system. We will use the notation $t_{ij}$ to denote $\lvert t_i - t_j\lvert$. The linear equation is
\begin{align}
\begin{bmatrix}
1 & \exp(-\beta t_{12}) & \exp(-\beta t_{13}) & \exp(-\beta t_{14})\\
\exp(-\beta t_{12}) & 1 & \exp(-\beta t_{23}) & \exp(-\beta t_{24})\\
\exp(-\beta t_{13}) & \exp(-\beta t_{23}) & 1 & \exp(-\beta t_{34})\\
\exp(-\beta t_{14}) & \exp(-\beta t_{24}) & \exp(-\beta t_{34}) & 1\\
\end{bmatrix}
\begin{bmatrix}
x_1\\x_2\\x_3\\x_4
\end{bmatrix}
=
\begin{bmatrix}
b_1\\b_2\\b_3\\b_4
\end{bmatrix}
\end{align}
Now lets introduce the additional variables as follows:
\begin{align}
r_4 & = x_4\\
r_3 & = x_3 + \exp(-\beta t_{34})r_4\\
r_2 & = x_2 + \exp(-\beta t_{23})r_3
\end{align}
\begin{align}
l_2 & = x_1\exp(-\beta t_{12})\\
l_3 & = (x_2+l_2)\exp(-\beta t_{23})\\
l_4 & = (x_3+l_3)\exp(-\beta t_{34})
\end{align}
The equations then become
\begin{align}
x_1 + \exp(-\beta t_{12}) r_2 & = b_1\\
l_2 + x_2 + \exp(-\beta t_{23}) r_3 & = b_2\\
l_3 + x_3 + \exp(-\beta t_{24}) r_4 & = b_3\\
l_4 + x_4 & = b_4
\end{align}
Embedding this in an extended sparse matrix, we obtain
\begin{align}
\begin{bmatrix}
1 & \exp(-\beta t_{12}) & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
\exp(-\beta t_{12}) & 0 & -1 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
0 & -1 & 0 & 1 & \exp(-\beta t_{23}) & 0 & 0 & 0 & 0 & 0\\
0 & 0 & 1 & 1 & \exp(-\beta t_{23}) & 0 & 0 & 0 & 0 & 0\\
0 & 0 & \exp(-\beta t_{23}) & \exp(-\beta t_{23}) & 0 & -1 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & -1 & 0 & 1 & \exp(-\beta t_{34}) & 0 & 0\\
0 & 0 & 0 & 0 & 0 & 1 & 1 & \exp(-\beta t_{34}) & 0 & 0\\
0 & 0 & 0 & 0 & 0 & \exp(-\beta t_{34}) & \exp(-\beta t_{34}) & 0 & -1 & 0\\
0 & 0 & 0 & 0 & 0 & 0 & 0 & -1 & 0 & 1\\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
\end{bmatrix}
\begin{bmatrix}
x_1\\ r_2\\ l_1\\ x_2\\ r_3\\ l_2\\ x_3\\ r_4\\ l_3\\ x_4
\end{bmatrix}
=
\begin{bmatrix}
b_1\\ 0\\ 0\\ b_2\\ 0\\ 0\\ b_3\\ 0\\ 0\\ b_4
\end{bmatrix}
\end{align}

@@

\newpage
% \bibliography{biblio}
% \bibliographystyle{unsrt}

\begin{thebibliography}{10}

@@

\bibitem{rybicki1995class}
George~B Rybicki and William~H Press.
\newblock Class of fast methods for processing irregularly sampled or otherwise
  inhomogeneous one-dimensional data.
\newblock {\em Physical review letters}, 74(7):1060, 1995.
